---
title: "Data_Analysis_Visulisation_R"
author: "armin"
date: "9/4/2020"
output: html_document
---

Task A: Data on the Indigenous Australian Population

A.1 Investigate the Distribution of Indigenous Australians 

1) Use R to read, wrangle and analyze the data in Data1. Make sure you describe any
complications you encounter and the steps you take when answering the following questions.


```{r}
# install packages
library(tidyverse)
library(ggplot2)

# reading in csv data from local file
data_1 <- read.csv("FIT5145_S2_2020_A1_Data1_IndigAusPopData_byregion.csv")

# see data types and basic data overview
glimpse(data_1)

## basic wrangling steps
# Age data was faulty so had to adjust format
# change Oct-14 to a proper age = 10-14
data_1$Age[data_1$Age == "Oct-14"] <- "10 - 14"

# change 5-Sep to a proper age = 5-9
data_1$Age[data_1$Age == "5-Sep"] <- "5 - 9"

# set column Age for character to factor
data_1$Age <- as.factor(data_1$Age)

```

- when going through data frame data_1 I realized that some of the Age values were not formatted correctly
  and were saved as a date and not a usable Age value
  
  example: "Oct-14" "5-Sep" were obviously not correct age values and were thus transformed to
  "10 - 14" and "5 - 9"

NEW "STATE" COLUMN FOR DATA_1
This state column is needed for tasks further down the line where we analyze
data on a state level. Hence, we need to associate a state with each territory. 
```{r}
# adding extra column to data frame data_1 showing the state associated with each region

# view data frame
glimpse(data_1)

# set column state from character to factor
data_1$Indigenous.region <- as.factor(data_1$Indigenous.region)

# get all unique regions
levels(data_1$Indigenous.region)

# add a column named state
data_1['State'] <- "NA"

# manually add the matching state to each region/city/territory 
data_1$State[data_1$Indigenous.region == "ACT"] <- "ACT"
data_1$State[data_1$Indigenous.region == "Adelaide"] <- "SA"
data_1$State[data_1$Indigenous.region == "Alice Springs"] <- "NT"
data_1$State[data_1$Indigenous.region == "Apatula"] <- "NT"
data_1$State[data_1$Indigenous.region == "Brisbane"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Broome"] <- "WA"
data_1$State[data_1$Indigenous.region == "Cairns - Atherton"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Cape York"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Darwin"] <- "NT"
data_1$State[data_1$Indigenous.region == "Dubbo"] <- "NSW"
data_1$State[data_1$Indigenous.region == "Geraldton"] <- "WA"
data_1$State[data_1$Indigenous.region == "Jabiru - Tiwi"] <- "NT"
data_1$State[data_1$Indigenous.region == "Kalgoorlie"] <- "WA"
data_1$State[data_1$Indigenous.region == "Katherine"] <- "NT"
data_1$State[data_1$Indigenous.region == "Kununurra"] <- "WA"
data_1$State[data_1$Indigenous.region == "Melbourne"] <- "Vic"
data_1$State[data_1$Indigenous.region == "Mount Isa"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Nhulunbuy"] <- "NT"
data_1$State[data_1$Indigenous.region == "North-Eastern NSW"] <- "NSW"
data_1$State[data_1$Indigenous.region == "North-Western NSW"] <- "NSW"
data_1$State[data_1$Indigenous.region == "NSW Central and North Coast"] <- "NSW"
data_1$State[data_1$Indigenous.region == "Perth"] <- "WA"
data_1$State[data_1$Indigenous.region == "Port Augusta"] <- "SA"
data_1$State[data_1$Indigenous.region == "Port Lincoln - Ceduna"] <- "SA"
data_1$State[data_1$Indigenous.region == "Riverina - Orange"] <- "WA"
data_1$State[data_1$Indigenous.region == "Rockhampton"] <- "QLD"
data_1$State[data_1$Indigenous.region == "South Hedland"] <- "WA"
data_1$State[data_1$Indigenous.region == "South-Eastern NSW"] <- "NSW"
data_1$State[data_1$Indigenous.region == "South-Western WA"] <- "WA"
data_1$State[data_1$Indigenous.region == "Sydney - Wollongong"] <- "NSW"
data_1$State[data_1$Indigenous.region == "Tasmania"] <- "Tas"
data_1$State[data_1$Indigenous.region == "Tennant Creek"] <- "NT"
data_1$State[data_1$Indigenous.region == "Toowoomba - Roma"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Torres Strait"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Townsville - Mackay"] <- "QLD"
data_1$State[data_1$Indigenous.region == "Victoria exc. Melbourne"] <- "Vic"
data_1$State[data_1$Indigenous.region == "West Kimberley"] <- "WA"

# now all territories/ cities have a state associated with them
```


a. What regions have the maximum and minimum total Indigenous populations in 2016 and
2031?

2016: Max: "NSW Central and North Coast"
      Min: "Port Lincoln - Ceduna"

2031: Max: "Brisbane"
      Min: "Port Lincoln - Ceduna"

```{r}
# aggregate data by region and time
max_min_indig_pop <- aggregate(x=data_1$Value,by=list(data_1$Indigenous.region,data_1$TIME),FUN=sum)

# rename the columns of data frame
names(max_min_indig_pop) <- c('region','year','count')

# sub select based on the required criteria 
# first get the wanted year (2016), then the population count
# after that use which.max to get the maximum value and then retrieve the value associated with it
max_pop_region_2016 <- max_min_indig_pop$region[which.max(max_min_indig_pop$count[max_min_indig_pop$year== 2016])]

# now do the same but with which.min
min_pop_region_2016 <- max_min_indig_pop$region[which.min(max_min_indig_pop$count[max_min_indig_pop$year== 2016])]

# print answer
max_pop_region_2016 
# print answer
min_pop_region_2016 

# sub select based on the required criteria
# first get the wanted year (2031), then the population count
# after that use which.max to get the maximum value and then retrieve the value associated with it
max_pop_region_2031 <- max_min_indig_pop$region[which.max(max_min_indig_pop$count[max_min_indig_pop$year== 2031])]
# now do the same but with which.min
min_pop_region_2031 <- max_min_indig_pop$region[which.min(max_min_indig_pop$count[max_min_indig_pop$year== 2031])]

# print answer
max_pop_region_2031
# print answer
min_pop_region_2031

```


b. What region/s have the maximum and minimum growth or decay rates of their total
Indigenous population between 2016 and 2031?

Max growth answer: "ACT"

Min growth/decay answer: "Katherine"

```{r}
# aggregate data by region and time
max_min_indig_pop <- aggregate(x=data_1$Value,by=list(data_1$Indigenous.region,data_1$TIME),FUN=sum)

# rename the columns of data frame max_min_indig_pop
names(max_min_indig_pop) <- c('region','year','count')

# get the maximum growth rate from 2016 to 2031
# first, set the year to 2016 and get the population count, do the same thing for 2031
# second, subtract them from another and divide by the 2016 value
# finally, get the max value of that percentage calculation and display the region
max_growth_reg <- max_min_indig_pop$region[which.max((((max_min_indig_pop$count[max_min_indig_pop$year== 2031]) - (max_min_indig_pop$count[max_min_indig_pop$year== 2016]))/(max_min_indig_pop$count[max_min_indig_pop$year== 2016])))]

# get the minimum growth rate from 2016 to 2031
# first, set the year to 2016 and get the population count, do the same thing for 2031
# second, subtract them from another and divide by the 2016 value
# finally, get the min value of that percentage calculation and display the region
min_growth_reg <- max_min_indig_pop$region[which.min((((max_min_indig_pop$count[max_min_indig_pop$year== 2031]) - (max_min_indig_pop$count[max_min_indig_pop$year== 2016]))/(max_min_indig_pop$count[max_min_indig_pop$year== 2016])))]

# print out max_growth_reg
max_growth_reg

# print out min_growth_reg
min_growth_reg
```


c. Plot and describe the growth or decay of the total Indigenous populations for the capitals of
the 8 state/territories across all time periods.

Answer: see below code and graph of sub task c 

```{r}
# get all unique locations 
# needed for manually extracting all the capital city names
unique(data_1$Indigenous.region)

# all the capital cities that are in the $Indigenous.region column 
capital_cities <- list("ACT","Adelaide","Alice Springs","Brisbane","Darwin","Melbourne","Sydney - Wollongong","Tasmania")

# only save the regions which are capital cities
# cities that are in data_1$Indigenous.region %in% as well as in capital_cities
indig_pop <- data_1[data_1$Indigenous.region %in% capital_cities,]

# aggregate/group by time and region based on population count
indig_pop <- aggregate(indig_pop$Value, FUN = sum, by = list(indig_pop$TIME, indig_pop$Indigenous.region))

# rename the columns in the data frame
names(indig_pop) <- c('year','state','count')

# turn each unique state into a factor
indig_pop$state <- as.factor(indig_pop$state)

levels(indig_pop$state)

# create list named dat, used for data frames that we create
dat <-list()

# now we calculate all the growth rates for each capital over time
# loop through all years from 2016 top 2031
for (i in unique(indig_pop$year)){
    
    # calculate the rate of change in population each year, just as done in part b
    rate <- (indig_pop$count[indig_pop$year==i+1]-indig_pop$count[indig_pop$year==i])/indig_pop$count[indig_pop$year==i]
    # add a column named year to show the time period observed
    year <- paste(i, '-',i+1)
    # part of the newly created year column 
    state <- indig_pop$state[indig_pop$year==i]
    # use of tibble function from tidyverse to create a new "data frame" containing rate of population change, state and time period
    df <- tibble(rate,state, year)
    # append the data frame df to list dat 
    dat[[i]]<-df
}
# NOTE: this for loop always throws a small error but still woks and serves its purpose


# concatenate all data frames in the list dat into one data frame called new_data_frame
new_data_frame <- do.call(rbind, dat)

# check out the new data frame with a growth rate column 
glimpse(new_data_frame)

# plot the data using ggplot
# geom_point(....) creates an empty cartisian plane
# x axis = year
# y axis = rate 
ggplot(data = new_data_frame) + 
  geom_point(mapping = aes(x = year, y = rate)) + 
  facet_wrap(~ state, nrow = 2)

# plot enables us to read off trends over time for each city
```

Graph Description:

Across all the eight capital cities population development no overarching pattern
can be observed. Thus, we will describe each city individually.

"ACT": indigenous population is stagnant/very slowly growing
"Adelaide": indigenous population grew from 2016-2025 but is plateauing after that 
"Alice Springs": indigenous population is continuously shrinking
"Brisbane": indigenous population grew from 2016-2025 but is shrinking back to 2016 levels
            past 2025
"Darwin": indigenous population is slowly decrease
"Melbourne": indigenous population is continuously growing
"Sydney - Wollongon": indigenous population is continuously growing
"Tasmania": indigenous population is continuously growing but growth seems to
            slow from 2025 onward


A.2 Investigate the ages of indigenous Australians

1. Which region has the highest percentage of children in its total 2016 population?
(children are < 15)

Answer: "Toowoomba - Roma"

```{r}
# extract the subset where year is 2016
data_2016 <- subset(data_1,TIME == "2016")

# extract the subset where age is under 15 (aka kids)
data_kids_2016 <- subset(data_2016,Age == "0 - 4" | Age ==   "10 - 14" | Age ==   "5 - 9")

# aggregate/group by region based on population count to get a new data frame
# for general population count per city 
indig_pop <- aggregate(data_2016$Value, FUN = sum, by = list(data_2016$Indigenous.region))

# aggregate/group by region based on kids count to get a new data frame
# for kids population count per city 
indig_pop_kids <- aggregate(data_kids_2016$Value, FUN = sum, by = list(data_kids_2016$Indigenous.region))

# rename the columns in the data frame
names(indig_pop) <- c('state','count')

# rename the columns in the data frame
names(indig_pop_kids) <- c('state','count')

# calculate the state with the highest percentage of children in its total 2016 population
max_kids_2016 <- indig_pop$state[which.max(indig_pop_kids$count / indig_pop$count)]

# print result
max_kids_2016

```


2. Using Data2 only, calculate and discuss which state or territory has the highest percentage of
children in its total 2006, 2016 and 2031 populations.

Answer: 

Sate with highest percentage of children in its total population:
2006 = "QLD"
2016 = "QLD"
2031 = "QLD"

```{r}
## basic wrangling steps take for data_2 data frame
# reading in csv data from local file
data_2 <- read.csv("FIT5145_S2_2020_A1_Data2_IndigAusPopData_bystate.csv")

# change Age data type to dbl from character
data_2$Age <- as.numeric(data_2$Age)

# removing all the commas with nothing in the entire data frame
# code logic of user: John Kane-2 was imitated
# from: https://r.789695.n4.nabble.com/help-with-replacing-all-comma-in-a-data-frame-with-a-dot-td846131.html
# Harvard Footnote citation:
#R.789695.n4.nabble.com. 2020. R Help - Help With Replacing All Comma In A Data
#Frame With A Dot. [online] Available at: 
# <https://r.789695.n4.nabble.com/help-with-replacing-all-comma-in-a-data-frame-with-a-dot-td846131.html>
#[Accessed 7 September 2020].
substi <- function(x) {sub(",","",x)}
data_2 <-  data.frame(apply(data_2, 2, substi ))

# check format and see if wrangling step worked
glimpse(data_2)

# change data type to dbl from character
data_2$X2006 <- as.numeric(data_2$X2006)

# change data type to dbl from character
data_2$X2016 <- as.numeric(data_2$X2016)

# change data type to dbl from character
data_2$X2031 <- as.numeric(data_2$X2031)

glimpse(data_2)

## create sub data frames for 2006,16 and 31
# create sub data frame for only 2006
data_2_2006 <- select(data_2, Age, State, X2006)

# create sub data frame for only 2016
data_2_2016 <- select(data_2, Age, State, X2016)

# create sub data frame for only 2031
data_2_2031 <- select(data_2, Age, State, X2031)

## filter out only kids 
# filter kids 2006
kids_2006 <- filter(data_2_2006, Age <15)

# filter kids 2006
kids_2016 <- filter(data_2_2016, Age <15)

# filter kids 2006
kids_2031 <- filter(data_2_2031, Age <15)

## aggregate kids count for each of the three years
# aggregate/group by region based on kids count
kids_pop_2006 <- aggregate(kids_2006$X2006, FUN = sum, by = list(kids_2006$State))

# aggregate/group by region based on kids count
kids_pop_2016 <- aggregate(kids_2016$X2016, FUN = sum, by = list(kids_2016$State))

# aggregate/group by region based on kids count
kids_pop_2031 <- aggregate(kids_2031$X2031, FUN = sum, by = list(kids_2031$State))

## aggregate entire population count for each of the three years
# aggregate/group by region based on pop count
all_pop_2006 <- aggregate(data_2_2006$X2006, FUN = sum, by = list(data_2_2006$State))

# aggregate/group by region based on pop count
all_pop_2016 <- aggregate(data_2_2016$X2016, FUN = sum, by = list(data_2_2016$State))

# aggregate/group by region based pop count
all_pop_2031 <- aggregate(data_2_2031$X2031, FUN = sum, by = list(data_2_2031$State))

## calculate the % of kids in population for each year

# get max kids % 2006
per_kid_2006 <- data.frame(kids_pop_2006$Group.1, kids_pop_2006$x/all_pop_2006$x)
# rename the columns in the data frame
names(per_kid_2006) <- c('state','count')
# find the state with the highest % of children in its population in the year 20
max_kids_2006 <- per_kid_2006$state[which.max(per_kid_2006$count)]
# print answer (state wit the largest percentage of kids in the population for 2006)
max_kids_2006

# get max kids % 2016
per_kid_2016 <- data.frame(kids_pop_2016$Group.1, kids_pop_2016$x/all_pop_2016$x)
# rename the columns in the data frame
names(per_kid_2016) <- c('state','count')
# find the state with the highest % of children in its population in the year 20
max_kids_2016 <- per_kid_2016$state[which.max(per_kid_2016$count)]
# print answer (state wit the largest percentage of kids in the population for 2016)
max_kids_2016

# get max kids % 2031
per_kid_2031 <- data.frame(kids_pop_2031$Group.1, kids_pop_2031$x/all_pop_2031$x)
# rename the columns in the data frame
names(per_kid_2031) <- c('state','count')
# find the state with the highest % of children in its population in the year 20
max_kids_2031 <- per_kid_2016$state[which.max(per_kid_2031$count)]
# print answer (state wit the largest percentage of kids in the population for 2031)
max_kids_2031


```


3. Use R to build a Motion Chart comparing the total Indigenous Australian population of each
region to the percentage of Indigenous Australian children in each state/territory. Use the
region populations calculated from Data1 and the child percentage values calculated
from Data2. The motion chart should show the population on the x-axis, the percentage on
the y-axis, the bubble size should depend on the population.

Answer: run the code and a tab with the motion chart will open, 
        IMPORTANT: enable "Flash" plugins in your browser for the chart to load! 

```{r}
# read in data_2 again from its csv format
data_2 <- read.csv("FIT5145_S2_2020_A1_Data2_IndigAusPopData_bystate.csv")


# like earlier we substitute all commas with nothing throughout the entire
# data frame
# Harvard Footnote citation:
#R.789695.n4.nabble.com. 2020. R Help - Help With Replacing All Comma In A Data
#Frame With A Dot. [online] Available at: 
# <https://r.789695.n4.nabble.com/help-with-replacing-all-comma-in-a-data-frame-with-a-dot-td846131.html>
#[Accessed 7 September 2020].
substi <- function(x) {sub(",","",x)}
data_2 <-  data.frame(apply(data_2, 2, substi ))

# renaming columns of data_2, since the years are a sequence we can apply the
# str_sub() function instead of manually writing out each year column name
names(data_2)<- c('Age', 'State', str_sub(names(data_2[, c(3:28)]), 2, -1))

# retrieving the children population, anyone under the age of 15, of each state
# and year
# %>%/piping operator is used to narrow down our sub selection and only work with
# the desired data
# use gather to pivot the table (change x and y axis)
children_pop_state <- data_2%>% gather(key = 'Year', value = 'Value', -Age, -State) %>% filter(as.numeric(Age)<15)
# turn the population values into numbers
children_pop_state$Value <- as.numeric(children_pop_state$Value)
# aggregate by both year and state and sum the population for each case
children_pop_state <- aggregate(children_pop_state$Value, by = list(children_pop_state$Year,children_pop_state$State), FUN = sum)
# change column names 
names(children_pop_state) <- c('Year', 'State', 'chlildren')

# retrieving total population of each state and year
# and use gather to pivot the table (change x and y axis)
total_pop_state <- data_2%>% gather(key = 'Year', value = 'Value', -Age, -State)
# aggregate by both year and state and sum the population for each case
total_pop_state<- aggregate(as.numeric(total_pop_state$Value), by = list(total_pop_state$Year,total_pop_state$State), FUN = sum)
# change column names 
names(total_pop_state) <- c('Year', 'State', 'Population')


# create a new data frame containing the children ratio of each state and then use
# for that purpose we had to merge the children_pop_state and the total_pop_state
# data frames
# merge based on year and state
children_ratio <- merge(children_pop_state, total_pop_state, on = c('Year', 'State'))
# divide the number of children by the number of the total population
children_ratio$Ratio <- children_ratio$chlildren/children_ratio$Population
# now only keep the year, state, and ratio
children_ratio%>% select(Year, State, Ratio)

# retrieving total population of the region 
region_pop <- aggregate(data_1$Value, by = list(data_1$TIME, data_1$Indigenous.region, data_1$State), FUN =sum)
# change column names 
names(region_pop) <- c('Year', 'Region', 'State', 'Population')

# now merge the data_1 and data_2 information into one data frame to get the 
# kids ratio for each region and state over time
children_ratio_to_region_pop<- merge(children_ratio%>% select(Year, State, Ratio), region_pop, on = c('Year', 'State'))
# change years to a numeric data type
children_ratio_to_region_pop$Year <- as.numeric(children_ratio_to_region_pop$Year)


# now plot the motion chart using the above created data frames
# need the package googleVis for that to work
install.packages('googleVis')
library(googleVis)

# define what should be displayed in the motion chart
# xvar = 'Population'     --> x axis
# yvar = 'Ratio'          --> y axis
# colorvar = 'State'      --> color of bubbles
# sizevar = 'Population'  --> size of bubbles
# timevar = 'Year'        --> movable x axis 
# idvar = 'Region'        --> sub categorie of each bubble
motion <- gvisMotionChart(children_ratio_to_region_pop, 
                          xvar = 'Population', 
                          yvar = 'Ratio', 
                          colorvar = 'State', 
                          sizevar = 'Population', 
                          timevar = 'Year', 
                          idvar = 'Region')
# now plot the actual chart
# a tab will open automatically 
plot(motion)

```

4. Using the Motion Chart, answer the following questions:

a. Which region’s population overtakes that of another region in the same
state/territory? In which year/s does this happen?

Answer:In the Northern Territory Apatula overtakes Katherine in the year 2021

b. Is there generally a relationship between the Indigenous Australian
population size and percentage of children in the population? If so, what kind
of relationship? Explain your answer.

Answer: The relationship appears to be linear to an extent. As the population size
increases the % of children increases. This might be due to the fact that a lot
of young families move to larger cities since smaller and more rural communities
lack opportunities and other infrastructure. I would expect population size and
percentage of children in the population to be slightly positively correlated.

However, in order to find out the precise relationship between the Indigenous
Australian population size and percentage of children in the population one would
have to conduct a linear regression between the two factors to determine their\
actual correlation.

c. Colour is commonly used in data visualisation to help understand data. Which
aspect of this data would you use colour for in your plot and why?

Answer: I used color for the different states since it makes it easier to
see trends between states over time that way. Color is commonly used in motion 
charts to compare similar groups such as countries and states.

d. Are there any other interesting things you notice in the data or any changes
you would recommend for the Motion Chart?

Answer:I wound readjust the scale of the chart to make changes over time more
noticeable. Since most of the population size and percentage of children in the
population changes were gradual, it would make more sense to show the
movements more in detail.


Task B: Exploratory Analysis on Australian Immunization rates

B1. Values and Variables

First we load the data and do basic wrangeling steps again to use the data from
data_3 for further analysis.
```{r}
# install packages
library(tidyverse)
library(ggplot2)

# reading in csv data from local file
data_3 <- read.csv("FIT5145_S2_2020_A1_Data3_AusImmunisationData.csv")

# get idea of data
glimpse(data_3)

## basic wrangling 

# notice the occurrence of unused NA columns X, X.1, X.2, X.3 and remove them 
# drop the 13-16th column 
data_3 <- data_3[ -c(13:16) ]

# removing all the commas with nothing in the entire data frame
# code logic of user: John Kane-2 was imitated
# from: https://r.789695.n4.nabble.com/help-with-replacing-all-comma-in-a-data-frame-with-a-dot-td846131.html
substi <- function(x) {sub(",","",x)}
data_3 <-  data.frame(apply(data_3, 2, substi ))

```

COLUMN FOR DATA_3
```{r}
## adding an extra column to data frame data_3 showing the state associated

# view data frame
glimpse(data_3)

# turn column PHN code into a factor 
data_3$PHN.area.name <- as.factor(data_3$PHN.area.name)

levels(data_3$PHN.area.name)


# add a column named state to data_3
data_3['State'] <- "NA"

# manually associate each region with a state
data_3$State[data_3$PHN.area.name == "Adelaide"] <- "SA"
data_3$State[data_3$PHN.area.name == "Australian Capital Territory"] <- "ACT"
data_3$State[data_3$PHN.area.name == "Brisbane North"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Brisbane South"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Central and Eastern Sydney"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Central Queensland Wide Bay and Sunshine Coast"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Country SA"] <- "SA"
data_3$State[data_3$PHN.area.name == "Country WA"] <- "WA"
data_3$State[data_3$PHN.area.name == "Darling Downs and West Moreton"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Eastern Melbourne"] <- "Vic"
data_3$State[data_3$PHN.area.name == "Gippsland"] <- "Vic"
data_3$State[data_3$PHN.area.name == "Gold Coast"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Hunter New England and Central Coast"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Murray"] <- "Vic"
data_3$State[data_3$PHN.area.name == "Murrumbidgee"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Nepean Blue Mountains"] <- "NSW"
data_3$State[data_3$PHN.area.name == "North Coast"] <- "NSW"
data_3$State[data_3$PHN.area.name == "North Western Melbourne"] <- "Vic"
data_3$State[data_3$PHN.area.name == "Northern Queensland"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Northern Sydney"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Northern Territory"] <- "NT"
data_3$State[data_3$PHN.area.name == "Perth North"] <- "WA"
data_3$State[data_3$PHN.area.name == "Perth South"] <- "WA"
data_3$State[data_3$PHN.area.name == "South Eastern Melbourne"] <- "Vic"
data_3$State[data_3$PHN.area.name == "South Eastern NSW"] <- "NSW"
data_3$State[data_3$PHN.area.name == "South Western Sydney"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Tasmania"] <- "Tas"
data_3$State[data_3$PHN.area.name == "Western NSW"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Western Queensland"] <- "QLD"
data_3$State[data_3$PHN.area.name == "Western Sydney"] <- "NSW"
data_3$State[data_3$PHN.area.name == "Western Victoria"] <- "Vic"


# now each region is associated with a state
# check change
glimpse(data_3)
```



1. How many PHN areas does the data cover?

Answer: 31

```{r}
# turn column PHN code into a factor 
data_3$PHN.code <- as.factor(data_3$PHN.code)

# get length of levels to know the number of area codes
num_PHN_areas <- length(levels(data_3$PHN.code))

# print answer
num_PHN_areas

```


2. What are the possible values for 'PHN code'?

Answer: There are 31 possible values which are:

"PHN101" "PHN102" "PHN103" "PHN104" "PHN105" "PHN106" "PHN107" "PHN108" "PHN109"
"PHN110" "PHN201" "PHN202" "PHN203" "PHN204" "PHN205" "PHN206" "PHN301" "PHN302"
"PHN303" "PHN304" "PHN305" "PHN306" "PHN307" "PHN401" "PHN402" "PHN501" "PHN502"
"PHN503" "PHN601" "PHN701" "PHN801"
```{r}
# get all PHN code levels as a vector
all_PHN_areas <- c(levels(data_3$PHN.code))

# print answer 
all_PHN_areas
```


3. For each row, calculate the percentage of Australian children that are fully immunised (this is
the immunisation rate). What are the average, maximum and minimum immunisation rates?
Calculate the same for the group that are Indigenous Australian children. Do all of those
values seem statistically reasonable to you?

Answers:

The main obstacle in this question were the missing values. In order to calculate
the mean we had to leave these value out.

QUESTION: Calculate the percentage of Australian children that are fully immunized (this is
the immunization rate): 

All immunization rates are saved in the column "data_3$immunize_rate"

All population immunization rates:
average immunization rate: 0.9206154
max immunization rate:     0.9615255
min immunization rate:     0.8609929

QUESTION: Calculate the percentage of Indigenous Australian children that are
fully immunized (this is the immunization rate): 

All immunization rates for Indigenous Australian children are saved in the
column "data_3$immunize_rate_indig"

Indigenous Australian immunization rates:
average immunization rate: 0.8992436 (when skipping NA values, using na.rm = TRUE)
max immunization rate:     0.9768977
min immunization rate:     0.7377049  

QUESTION: Do all of those values seem statistically reasonable to you?

All the values seem reasonable after removing NA values. Given that a lot of Indigenous
Australians have worse access to health care it is understandable that their mean
immunization rate would be lower than of the entire Australian Population.
Moreover, using the same set of assumptions it seems reasonable that the minimum
immunization rate for Indigenous Australians kids would be lower than of the
entire Australian Population.


```{r}

# change data type to dbl from character
data_3$Number.of.registered.children <- as.numeric(data_3$Number.of.registered.children)

# change data type to dbl from character
data_3$Number.fully.immunised <- as.numeric(data_3$Number.fully.immunised)

# change data type to dbl from character
data_3$Number.of.registered.IndigAus..children <- as.numeric(data_3$Number.of.registered.IndigAus..children)

# change data type to dbl from character
data_3$Number.IndigAus.fully.immunised <- as.numeric(data_3$Number.IndigAus.fully.immunised)


# loop through all the rows of data frame data_3
# calculate the desired immunization rates
for (x in 1:nrow(data_3)){
    
    # calculate the immunization rate for all pop
    immu_rate <- (data_3$Number.fully.immunised/data_3$Number.of.registered.children)
    # add a column named immunize_rate
    data_3['immunize_rate'] <- immu_rate
    
    # calculate the immunization rate for indigenous pop
    immu_rate_indig <- (data_3$Number.IndigAus.fully.immunised/data_3$Number.of.registered.IndigAus..children)
    # add a column named immunize_rate_indig
    data_3['immunize_rate_indig'] <- immu_rate_indig
    
}

## get immunization rate for all population
# get average immunization rate of all rows in data_3
avg_immu_rate <- mean(data_3$immunize_rate)

# print answer
avg_immu_rate 

# get max immunization rate of all rows in data_3
max_immu_rate <- data_3$immunize_rate[which.max(data_3$immunize_rate)]

# print answer
max_immu_rate 

# get max immunization rate of all rows in data_3
min_immu_rate <- data_3$immunize_rate[which.min(data_3$immunize_rate)]

# print answer
min_immu_rate 

## get immunization rate for indigenous population
# get average immunization rate of all rows in data_3

# Problem:
# first, when I tried to calculate the mean value of data_3$immunize_rate_indig
# I got NA, this is due to missing values in the column containing NA
# thus, in order to calculate the mean we need to skip the missing values
avg_immu_rate_indig <- mean(data_3$immunize_rate_indig, na.rm = TRUE)

# print answer
avg_immu_rate_indig 

# get max immunization rate of all rows in data_3
max_immu_rate_indig <- data_3$immunize_rate_indig[which.max(data_3$immunize_rate_indig)]

# print answer
max_immu_rate_indig 

# get max immunization rate of all rows in data_3
min_immu_rate_indig <- data_3$immunize_rate_indig[which.min(data_3$immunize_rate_indig)]

# print answer
min_immu_rate_indig 

```



B2. Variation in rates over Time, Age and Location

Generate boxplots (or other plots) of the immunisation rates versus year and age
to answer the following questions:

1. Have the immunisation rates improved over time? Are the median immunization
rates increasing, decreasing or staying the same?

Answer: As the scatter plot showcased median immunization rates decreased from
2013-2015 but increased again from 2016 onward.

```{r}

# aggregate/group by time and get median immunization rates
immu_over_time <- aggregate(data_3$immunize_rate, FUN = median, by = list(data_3$Reporting.Year))
# change column names
names(immu_over_time) <- c("year","immunization_rate")

# Basic scatter plot using ggplot to display the mean immunization rates over time
ggplot(immu_over_time, aes(x=year, y=immunization_rate)) + geom_point()


```


2. How do the immunisation rates vary with the age of the child?

Answer: By aggregating the data_3 data frame by age and plotting the mean immunization
rate it can be observed that immunization rates don't vary largely largely based
on age. However, the relatively lowest immunization rate can be seen in the age
group of 2 year old's, while the age groups of 1 and 5 year old's had marginally
higher immunization rates.

```{r}
# check relevant data
glimpse(data_3)
# aggregate/group by age and get mean immunization rates
immu_by_age <- aggregate(data_3$immunize_rate, FUN = mean, by = list(data_3$Age.group ))

# change column name
names(immu_by_age) <- c("age","immunization_rate")

# Basic scatter plot using ggplot
ggplot(immu_by_age, aes(x=age, y=immunization_rate)) + geom_point()
```

Generate boxplots (or other plots) of the immunisation rates versus locations
and answer the following questions:

3. What is the median rate per state/territory?

Answers:

State       Median rate
ACT	        0.9301651			
NSW	        0.9224214			
NT	        0.9140182			
QLD	        0.9239524			
SA	        0.9166372			
Tas	        0.9271472			
Vic	        0.9261011			
WA	        0.9077603	

```{r}
glimpse(data_3)
# aggregate/group by age and get mean immunization rates
immu_by_region <- aggregate(data_3$immunize_rate, FUN = median, by = list(data_3$State))

# change column name
names(immu_by_region) <- c("region","immunization_rate")

# print answer
immu_by_region

# Basic scatter plot using ggplot
ggplot(immu_by_region, aes(x=immunization_rate, y=region)) + geom_point()


```


4. Which states or territories seem most consistent in their immunization rates?

Answer: 
Territories:
After plotting all territories over time in consideration of
their mean immunisation rates it can be observed that territories such as ACT,
all of Brisbane, and all of Perth have consistent immunization rates over time.
On the other hand, territories such as Western Sydney, Northern Queensland,
Eastern Queensland, and Tasmania have had inconsistent immunization rates over
time.
This might be due to these regions having lower median household incomes and less
access to health care infrastructure.

States:
When looking at all states both ACT and WA have the most consistent mean
immunization rates over time. All the other states experience fluctuations in
their immunization rates of around 2-4% over the years.


```{r}

## observe mean immunization rates per territory 
# aggregate/group by age and territory and get mean immunization rates
immu_by_region_year <- aggregate(data_3$immunize_rate, FUN = median, by = list(data_3$Reporting.Year,data_3$PHN.area.name ))

# change column name
names(immu_by_region_year) <- c("year","region","immunization_rate")

# print data frame
immu_by_region_year

# plot immunization rates over time per territory/region
ggplot(data = immu_by_region_year) + 
  geom_point(mapping = aes(x = immunization_rate, y = year)) + 
  facet_wrap(~ region, nrow = 4)

## observe mean immunization rates per state
# aggregate/group by age and state and get mean immunization rates
immu_by_state_year <- aggregate(data_3$immunize_rate, FUN = median, by = list(data_3$Reporting.Year,data_3$State))


# change column name
names(immu_by_state_year) <- c("year","state","immunization_rate")

# plot immunization rates over time per state
ggplot(data = immu_by_state_year) + 
  geom_point(mapping = aes(x = immunization_rate, y = year)) + 
  facet_wrap(~ state, nrow = 4)

# Now there are two graphs: one for all territories and one for all states
```


